The \dealii Library, Version 9.1

### 2019-06-02 9.1/paper.tex
Write description of AD improvements

The major changes of this release are:
%
\begin{itemize}
  \item Improved support for automatic and symbolic differentiation (see
    Section~\ref{subsec:ad}),
  \item Full support for $hp$ adaptivity in parallel computations (see
    Section~\ref{subsec:hp}),
  \item Interfaces to the HDF5 file format and libraries (see
    Section~\ref{subsec:hdf5}),
  \item Significantly extended GPU support (see Section~\ref{subsec:gpu}),
  \item Parallel geometric multigrid (GMG) improvements (see
    \cite{ClevengerHeisterKanschatKronbichler2019} and
    Section~\ref{subsec:gmg}),
  \item Four new tutorial programs (step-61, step-62, step-63, step-64),
    as well as one new code gallery program (see
    Section~\ref{subsec:steps}).
\end{itemize}
%
The major changes are discussed in detail in Section~\ref{sec:major}. There
are a number of other noteworthy changes in the current \dealii{} release
that we briefly outline in the remainder of this section:
%
\begin{itemize}
\item
  The release contains bug fixes and performance improvements of the
  matrix-free framework and related geometric multigrid solvers. In
  particular, the implementation of the Chebyshev iteration, an often used
  smoother in the matrix-free context, has been revised to reduce vector
  accesses. This speeds up matrix-free multigrid solvers by up to 10-15\%
  on geometries with affine (parallelogram and parallelpiped) cells, and up
  to 5\% on geometries with cells bounded by curved edges and faces.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Improved support for automatic and symbolic differentiation}
\label{subsec:ad}

>>>>

The major changes of this release are:
%
\begin{itemize}
  \item Improved support for automatic differentiation (see
    Section~\ref{subsec:ad}),
  \item Full support for $hp$ adaptivity in parallel computations (see
    Section~\ref{subsec:hp}),
  \item Interfaces to the HDF5 file format and libraries (see
    Section~\ref{subsec:hdf5}),
  \item Significantly extended GPU support (see Section~\ref{subsec:gpu}),
  \item Parallel geometric multigrid (GMG) improvements (see
    \cite{ClevengerHeisterKanschatKronbichler2019} and
    Section~\ref{subsec:gmg}),
  \item Four new tutorial programs (step-61, step-62, step-63, step-64),
    as well as one new code gallery program (see
    Section~\ref{subsec:steps}).
\end{itemize}
%
The major changes are discussed in detail in Section~\ref{sec:major}. There
are a number of other noteworthy changes in the current \dealii{} release
that we briefly outline in the remainder of this section:
%
\begin{itemize}
\item
  The release contains bug fixes and performance improvements of the
  matrix-free framework and related geometric multigrid solvers. In
  particular, the implementation of the Chebyshev iteration, an often used
  smoother in the matrix-free context, has been revised to reduce vector
  accesses. This speeds up matrix-free multigrid solvers by up to 10-15\%
  on geometries with affine (parallelogram and parallelpiped) cells, and up
  to 5\% on geometries with cells bounded by curved edges and faces.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Improved support for automatic differentiation}
\label{subsec:ad}

In the previous release, numerous classes that are used to assemble linear systems
and right hand sides, as well those used to define constitutive laws, were given 
full support for ``white-listed'' automatically differentiable (AD) number types 
from the ADOL-C and Sacado libraries.
In the current release we have provided a unified interface to these
AD libraries focussing two specific use contexts, namely
\begin{enumerate}
\item the construction and linearization of finite element residuals, and
\item the construction and linearization of constitutive model kinetic variables.
\end{enumerate}

In the first context, the finite element degrees of freedom are considered the
independent variables. From these primitives, the \texttt{EnergyFunctional} helper
class in the namespace \texttt{Differentiation::AD} may be used to compute both the 
residual and its linearization by directly defining the contribution to the 
(twice differentiated) scalar total energy functional from each cell. Similarly, 
the \texttt{ResidualLinearization} class requires the (once-differentiated) finite 
element residual to be defined on a per cell basis,  and this contribution is 
automatically linearized.

The second context aims directly at constitutive model formulations, and serves to
compute the directional derivatives of components of (multi-field) constitutive laws
with respect to the scalar, vector, tensor and symmetric tensor fields in terms 
of which they are parameterized. The \texttt{ScalarFunction} class may be used to 
define a scalar function (such as strain energy function) that may be twice 
differentiated, while the \texttt{VectorFunction} may be used to define a vector
function (such as a set of kinematic fields) that may be differentiated once.
Since the total derivatives of all components are computed at once, these two helper 
classes provide an interface to retrieve each sub-component of the gradient and 
Hessian (for a \texttt{ScalarFunction}) or values and Jacobian (for a 
\texttt{VectorFunction}).

Although these aforementioned helper classes have been documented with a specific
use in mind, they remain generic and may (with a reinterpretation of the meaning of
the independent and dependent variables) be used for other purposes as well.
Furthermore, through the implementation of \texttt{TapedDrivers} and 
\texttt{TapelessDrivers} classes that interface with the active AD library, the 
generic helper classes hide library-dependent implementational details and facilitate 
switching between the supported libraries and AD number types based on the 
user's requirements.
